---
title: "Introduction to MetaboAnalystR"
subtitle: "Enrichment Analysis"
author: "Alex Sanchez-Pla"
date: "`r Sys.Date()`"
format:
    html: 
      toc: true
      toc-depth: 4
      number-sections: true
      embed-resources: true
    pdf:
      toc: true
      toc-depth: 3
      number-sections: true      
knit:
  quarto:
    chunk_options:
      echo: true
      cache: false
      prompt: false
      tidy: true
      comment: NA
      message: false
      warning: false
    knit_options:
      width: 75
# reference-location: margin
editor_options: 
  chunk_output_type: console
editor: 
  markdown: 
    wrap: 72
---

# Introduction to MetaboAnalyst

MetaboAnalystR comes with multiple vignetted ("tutorials") to illustrate how to perform distinct actions and processes.

The document [MetaboAnalystR Package](https://www.metaboanalyst.ca/docs/RTutorial.xhtml) provides an *overview* to the distinct vignettes and links to each of them.

The current document is a short introduction to Enrichment Analysis as can be done by MetaboAnalyst. It uses the results of the comparisons performed in the Statistical Analyses example.

# Enrichment Analysis Workflows

Below we will go over 2 use-cases to perform Enrichment Analysis, the first using as input a list of compounds, and the second as input a concentration table.

## Over representation analysis

### Preparation for ORA

The first step to ORA is to create a vector containing a list of compound names. 

The list will then be cross-referenced (CrossReferencing) against the MetaboAnalyst compound libraries (HMDB, PubChem, KEGG, etc.), and any compounds without a hit will have NA. 

This step may take long due to downloading of libraries if they do not already exist in your working directory.

```{r}
library(MetaboAnalystR)

## When input is a list

# Create vector consisting of compounds for enrichment analysis 
tmp.vec <- c("Acetoacetic acid", "Beta-Alanine", "Creatine", "Dimethylglycine", "Fumaric acid", "Glycine", "Homocysteine", "L-Cysteine", "L-Isolucine", "L-Phenylalanine", "L-Serine", "L-Threonine", "L-Tyrosine", "L-Valine", "Phenylpyruvic acid", "Propionic acid", "Pyruvic acid", "Sarcosine", "Arsenic", "Benzene", "Caffeic acid", "Cotinine", "Cadmium", "Lead", "Thiocyanate")

# Create mSetObj
mSet<-InitDataObjects("conc", "msetora", FALSE)
```

```{r}
#Set up mSetObj with the list of compounds
mSet<-Setup.MapData(mSet, tmp.vec);

# Cross reference list of compounds against libraries (hmdb, pubchem, chebi, kegg, metlin)
mSet<-CrossReferencing(mSet, "name");
```

To view the compound name map to identify any compounds within the uploaded list without hits…

```{r}
mSet$name.map 

```

### Enrichment Analysis


