
---
title: "Benchmarking Enrichment Packages"
subtitle: "ORA and MSEA for metabolomics using shared metabolite sets"
author: "Alex Sanchez"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: true
      smooth_scroll: true
    code_folding: hide
    self_contained: true
    number_sections: true
    theme: readable
knitr:
  opts_chunk:
    echo: true
    cache: false
    prompt: false
    tidy: true
    comment: NA
    message: false
    warning: false
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(stringr)
  library(readr)
  library(localEnrichment)
})


# Helper for nicer printing of tibbles
print_tbl <- function(x, n = 6) {
  print(dplyr::as_tibble(x), n = n)
}
```

# Introduction

Enrichment analysis aims to facilitate the biological interpretation of omics results by relating lists of features (genes, proteins, metabolites) to prior biological knowledge stored in databases (pathways, ontologies, chemical
classes, etc.).

In genomics, ORA and GSEA workflows are now fairly standardized. In metabolomics, however, tools are more fragmented and rely on different
identifier systems and pathway databases.

The goals of this document are:

1. **Quick-start:** show how to run basic over-representation analysis (ORA)  and metabolite set enrichment analysis (MSEA/GSEA-like) with several tools.
2. **Naïve benchmarking:** compare the results of these tools on the same datasets, using MetaboAnalyst Web output as a "practical reference"gold standard".

We will focus on:

- `enrichmet` (pathway enrichment, GSEA, centrality, plots)
- `localEnrichment` (our own infrastructure for metabolite sets and ORA/MSEA)
- `ChemRich` (which does not prohttps://raw.githubusercontent.com/barupal/ChemRICH/refs/heads/master/chemrich_minimum_analysis.R) 
- `MetaboAnalystR`, has become very hard to make it work so the examples will be run from its web site and downloaded.

Rather than exhaustively exploring each tool, we restrict ourselves to a reproducible workflow:

1. Use a **shared collection of metabolite sets** (`EnrichmentSet` objects).
2. For each dataset, derive:
   - a list of **significant metabolites** (*signature*),
   - a **background** tailored to that signature, using `filterEnrichmentSet()`.
3. Run ORA (and MSEA when possible) with the different tools on the same sets.


# Data and metabolite sets

Any pathway analysis requires two inputs:

- A list of features usually associated with a phenotype (e.g. metabolites differentially quantified between two groups). 
- A source of annotations for these features, that can be generically described as *Feature sets* or, as in this cas *Metabolite Sets*.

## The datasets

We use two datasets that have been preprocessed elsewhere (see file [Benchmarking-PrepareData.Rmd](Benchmarking-PrepareData.Rmd)).

1. The **Cachexia** dataset. This is a popular example used by MetaboAnalyst, where a comparison between affected and control patients is performed. It has the drawback that only  *differential*  metabolites are provided, which makes it only partially interesting for some types of analyses. We keep it for its popularity in MetaboAnalyst.

2. The **ST000291** dataset, downloaded from Metabolomics Workbench, and used in a fobitools use case. This is obtained by a nutritional intervention study where a comparion among consumption of Apple, Cranberries and Vaseline has been performed. In this case we have the full dataset and the results of comparing the groups using limma whih leads to around 40 differential metabolites. We focus only on the *Cranberry vs Baseline* comparison.

### Loadinhg datasets

The data for this use case has been prepared elsewhere and is available in the `data`directory.

```{r load-data}
# Cachexia: topTable + ID mapping
cachexia_map_path  <- "data/Cachexia_map.csv"
cachexia_tt_path   <- "data/Cachexia_topTable.Rda"

Cachexia_map <- read.csv(cachexia_map_path, stringsAsFactors = FALSE)
load(cachexia_tt_path)  # loads 'topCachexia'

cat("Cachexia_map:\n")
print_tbl(head(Cachexia_map))
cat("\ntopCachexia:\n")
print_tbl(head(topCachexia))

# ST000291: Cranberry vs Baseline topTable + ID mapping
st291_map_path   <- "data/MW_ST000291__map.csv"
st291_tt_path    <- "data/MW_ST000291-topCvsB.Rda"

MS_ST000291_map <- read.csv(st291_map_path, stringsAsFactors = FALSE)
load(st291_tt_path)  # loads 'topCranVSBase'

cat("\nMS_ST000291_map:\n")
print_tbl(head(MS_ST000291_map))
cat("\ntopCranVSBase:\n")
print_tbl(head(topCranVSBase))
```

## Metabolite Sets

We also use a **unified collection of metabolite sets**, previously constructed and saved as:

- `KEGGset_global`, `KEGGset_global_df`
- `KEGGset_study`,  `KEGGset_study_df`
- `ChemicalClassSet`, `ChemicalClassSet_df`
- `SMPDBset`, `SMPDBset_df`

All sets use **HMDB IDs** as feature identifiers but can be easily converted into **PubChemIDs** using the `metaboliteIDmapping` package.

### Load Metabolite Sets

Metabolite sets have been prepared elsewhere and saved as `EnrichmentSet` objects of the class defined in the `localEnrichment` package.

```{r load-data}
# Metabolite sets (EnrichmentSet collection)
mets_collection_path <- "databases/MetaboliteSets_Collection.Rda"
load(mets_collection_path)

cat("\nEnrichmentSet objects loaded:\n")
show(KEGGset_global);      cat("\n")
show(ChemicalClassSet);    cat("\n")
show(SMPDBset);            cat("\n")

cat("\nKEGGset_global_df:\n")
print_tbl(head(KEGGset_global_df))

cat("\nChemicalClassSet_df:\n")
print_tbl(head(ChemicalClassSet_df))

cat("\nSMPDBset_df:\n")
print_tbl(head(SMPDBset_df))

```


Notice that the data.frames contain the "set name" in the first column. For Pathway Analysis it may be preferable -depending on the tool used- to have the Set ids. This information is also contained in the `EnrichmentSet` object and can be easily extracted from there.

```{r}
KEGGset_global_df_ids <-as.MetaboliteSetDataFrame(KEGGset_global, id_type ="both")
head(KEGGset_global_df_ids)

SMPDBset_df_ids <-as.MetaboliteSetDataFrame(SMPDBset, id_type ="both")
head(SMPDBset_df_ids)

```


# Building signatures and backgrounds

Our aim is to derive **signatures** and **backgrounds** that are compatible with the `EnrichmentSet` objects, which use **HMDB IDs**.

We therefore:

1. Map each topTable to HMDB IDs.
2. Construct a vector of significant HMDB IDs (signature).
3. Use `filterEnrichmentSet(Eset, ids)` to restrict the metabolite sets to those
   that contain at least one of the signature metabolites.
4. Build a **background** as the union of all metabolites in the filtered sets.

## Cachexia: HMDB mapping and signature

```{r cachexia-hmdb}
library(tibble)

# Convert topCachexia to tibble with metabolite name
cachexia_tt <- topCachexia %>%
  rownames_to_column(var = "Metabolite") %>%
  left_join(Cachexia_map, by = c("Metabolite" = "Query"))

cat("Cachexia limma table with mapping:\n")
print_tbl(cachexia_tt)

# p-value column and threshold
p_col_cachexia <- "adj.P.Val"
alpha_cachexia <- 0.05

# Signature: HMDB IDs of significant metabolites
cachexia_sig_HMDB <- cachexia_tt %>%
  filter(!is.na(HMDB),
         !!rlang::sym(p_col_cachexia) < alpha_cachexia) %>%
  pull(HMDB) %>%
  unique()

length(cachexia_sig_HMDB)
cachexia_sig_HMDB
```

Note that in this Cachexia dataset we only have *differential* metabolites,
not the full list of measured features. A realistic background based purely
on the data is therefore impossible. Instead, we derive a *functional background* from the metabolite-set collection.

### Filter EnrichmentSets by signature (Cachexia)

We now use `filterEnrichmentSet()` to keep only sets that contain at
least one signature metabolite. This creates a **signature-induced subset**
of KEGG pathways (or other sets).

```{r cachexia-filter-es, eval=FALSE}
library(localEnrichment)  
# Filter KEGG global pathways to those that contain ≥1 Cachexia HMDB ID
KEGGset_cachexia <- filterEnrichmentSet(KEGGset_global, ids = cachexia_sig_HMDB)

show(KEGGset_cachexia)
```

To construct a background and a table suitable for `enrichmet()`, we need:

- The set names retained in `KEGGset_cachexia`.
- The HMDB members of each set.

The exact internal structure of `EnrichmentSet` may vary. A common pattern is
an S4 slot `sets` containing a named list of feature vectors. If this is the
case, we can do:

```{r cachexia-background, eval=FALSE}
# Example assuming an S4 slot 'sets' with a named list of HMDB vectors
slotNames(KEGGset_cachexia)

# Extract background as the union of all HMDB IDs in the filtered sets
sets_list_cachexia <- KEGGset_cachexia@sets
background_cachexia_HMDB <- unique(unlist(sets_list_cachexia))

length(background_cachexia_HMDB)

# Pathway ↔ metabolite table for enrichmet()
PathwayVsMetabolites_cachexia <- tibble::tibble(
  Pathway     = names(sets_list_cachexia),
  Metabolites = vapply(
    sets_list_cachexia,
    FUN = function(x) paste(x, collapse = ","),
    FUN.VALUE = character(1)
  )
)

print_tbl(head(PathwayVsMetabolites_cachexia))
```

> **Important:**  
> If your `EnrichmentSet` implementation uses *different* slot names or has
> accessor functions, adapt the lines using `KEGGset_cachexia@sets`
> accordingly. The idea is simply to retrieve:
> - a *named list* of HMDB vectors for each set, and
> - its names, which are the set identifiers.

If you prefer to avoid S4 internals, you can construct a filtered
`PathwayVsMetabolites` table using the accompanying `*_df` object and
the signature:

```{r cachexia-background-alt}
# Alternative: derive filtered KEGG df directly (without using slots)

# Expand global KEGG df to long format
kegg_long <- KEGGset_global_df %>%
  mutate(Metabolites = str_split(Metabolites, ",")) %>%
  unnest(Metabolites)

# Keep sets that contain ≥1 signature HMDB ID
selected_sets_cachexia <- kegg_long %>%
  filter(Metabolites %in% cachexia_sig_HMDB) %>%
  pull(set_name) %>%
  unique()

length(selected_sets_cachexia)

# Filter KEGG df to these sets
KEGGset_cachexia_df <- KEGGset_global_df %>%
  filter(set_name %in% selected_sets_cachexia)

# Functional background = union of all HMDB IDs in selected sets
background_cachexia_HMDB <- KEGGset_cachexia_df %>%
  mutate(Metabolites = str_split(Metabolites, ",")) %>%
  unnest(Metabolites) %>%
  pull(Metabolites) %>%
  unique()

length(background_cachexia_HMDB)
head(background_cachexia_HMDB)

# Data frame in the format expected by enrichmet()
PathwayVsMetabolites_cachexia <- KEGGset_cachexia_df %>%
  rename(Pathway = set_name)

print_tbl(head(PathwayVsMetabolites_cachexia))
```

In the remainder, we will use this **df-based version**, which is simpler and
does not depend on the internal structure of the `EnrichmentSet` object.


## ST000291: PubChem and HMDB mappings and signature

This dataset is annotated with PubChem identifiers, but the metaboliteSets available are based on HMDB.

The code below shows how to prepare a list of selected metabolites and background straight from the available information. 

```{r st291-Pbchem}
## ST000291: signature (PubChem) and background

# top table from limma
st291_tt <- topCranVSBase   # ja carregat anteriorment

# signature (adj.P.Val < 0.05)
sig_st291 <- st291_tt %>%
  filter(adj_pvalue  < 0.25) %>%
  pull(PubChemCID) %>%
  unique()

# background = ALL tested metabolites
bg_st291 <- st291_tt %>%
  pull(PubChemCID) %>%
  unique()

cat("Signature size:", length(sig_st291), "\n")
cat("Background size:", length(bg_st291), "\n")

```

These lists however cannot be directly used because of the way the MetaboliteSets are built.

Instead, we will convert the identifiers from PubChem to HMDB using the `metaboliteIDmapping` package. 


```{r}
############################################################
## ST000291 — Mapping PubChem → HMDB and Enrichment
############################################################

## 0. Load mapping table from metaboliteIDmapping
library(metaboliteIDmapping)

mapping_full <- metabolitesMapping   # tibble ~1M rows

# Keep only PubChem↔HMDB rows
mapping_pc_hmdb <- mapping_full %>%
  dplyr::filter(!is.na(CID), !is.na(HMDB)) %>%
  dplyr::select(CID, HMDB) %>%
  dplyr::distinct()

############################################################
## 1. Prepare limma topTable for ST000291
############################################################

# The object topCranVSBase is already loaded earlier in the Rmd
# It contains column: PubChemCID with numeric or character CIDs

bg_pubchem <- topCranVSBase$PubChemCID %>% unique()

sig_pubchem <- topCranVSBase %>%
  dplyr::filter(pvalue< 0.05) %>%
  dplyr::pull(PubChemCID) %>% unique()

cat("Signature PubChem:", length(sig_pubchem), "\n")
cat("Background PubChem:", length(bg_pubchem), "\n")


############################################################
## 2. PubChem → HMDB conversion (via metaboliteIDmapping)
############################################################

# Background HMDB
bg_hmdb <- mapping_pc_hmdb %>%
  dplyr::filter(CID %in% bg_pubchem) %>%
  dplyr::pull(HMDB) %>% unique()

# Signature HMDB
sig_hmdb <- mapping_pc_hmdb %>%
  dplyr::filter(CID %in% sig_pubchem) %>%
  dplyr::pull(HMDB) %>% unique()

cat("Signature HMDB:", length(sig_hmdb), "\n")
cat("Background HMDB:", length(bg_hmdb), "\n")


############################################################
## 3. Quantify ID loss
############################################################

mapped_ok  <- mapping_pc_hmdb %>%
  dplyr::filter(CID %in% bg_pubchem) %>%
  dplyr::pull(CID) %>% unique()

lost_pubchem <- setdiff(bg_pubchem, mapped_ok)

cat("Total PubChem in topTable:", length(bg_pubchem), "\n")
cat("Mapped PubChem→HMDB:", length(mapped_ok), "\n")
cat("Lost PubChem (no HMDB):", length(lost_pubchem), "\n")
cat("Loss %:", round(100 * length(lost_pubchem) / length(bg_pubchem), 1), "%\n")

# Optional: print the lost IDs
# lost_pubchem
```

### Filter EnrichmentSets by signature (ST000291)

We repeat the same logic for the ST000291 signature.

```{r st291-filter-es}
KEGG_st291_filt <- filterEnrichmentSet(KEGGset_global, bg_hmdb)
SMPDB_st291_filt <- filterEnrichmentSet(SMPDBset, bg_hmdb)
ChemClass_st291_filt <- filterEnrichmentSet(ChemicalClassSet, bg_hmdb)

############################################################
## Extract MetaboliteSet DataFrames from filtered EnrichmentSets
############################################################

# KEGG
KEGG_st291_df <- as.MetaboliteSetDataFrame(
  x = KEGG_st291_filt,
  id_type = "both",     # pots posar "id", "name", "both"
  id_sep  = ","         # separador dels metabolits
)

# SMPDB
SMPDB_st291_df <- as.MetaboliteSetDataFrame(
  x = SMPDB_st291_filt,
  id_type = "both",
  id_sep  = ","
)

# Chemical Classes
ChemClass_st291_df <- as.MetaboliteSetDataFrame(
  x = ChemClass_st291_filt,
  id_type = "both",
  id_sep  = ","
)

# Preview tables
cat("\nKEGG (filtered):\n")
print(head(KEGG_st291_df))

cat("\nSMPDB (filtered):\n")
print(head(SMPDB_st291_df))

cat("\nChemical Classes (filtered):\n")
print(head(ChemClass_st291_df))

```

We now have:

- `cachexia_sig_HMDB`, `background_cachexia_HMDB`,
  `PathwayVsMetabolites_cachexia`
- `sig_hmdb`, `bg_hmdb`,
  `KEGG_st291_df`, `SMPDB_st291_df`, `ChemClass_st291_df`

All in terms of HMDB IDs, consistent with the `EnrichmentSet` objects.

# Enrichment with enrichmet

We now use the `enrichmet()` function as defined in your package:

```r
enrichmet(
  inputMetabolites,
  PathwayVsMetabolites,
  example_data = NULL,
  top_n = 100,
  p_value_cutoff = 1,
  kegg_lookup = NULL,
  mapping_df = NULL,
  stitch_df = NULL,
  output_dir = NULL,
  save_excel = FALSE,
  analysis_type = c("enrichment", "gsea", "centrality", "network",
                    "heatmap", "membership", "interaction"),
  run_plots = TRUE
)
```

We will use:

- `"enrichment"` for ORA,
- `"gsea"` for MSEA/GSEA when we have a ranked list.

## ORA for Cachexia (enrichmet)

```{r enrichmet-cachexia-ora, eval=FALSE}
library(enrichmet)

results_cachexia_enrichmet <- enrichmet(
  inputMetabolites     = cachexia_sig_HMDB,
  PathwayVsMetabolites = PathwayVsMetabolites_cachexia,
  top_n                = 50,
  p_value_cutoff       = 0.05,
  analysis_type        = c("enrichment"),
  run_plots            = TRUE
)

cachexia_ora_enrichmet <- results_cachexia_enrichmet$pathway_enrichment_results
print_tbl(head(cachexia_ora_enrichmet))

# Optional plots:
# results_cachexia_enrichmet$pathway_plot
# results_cachexia_enrichmet$impact_plot
```

## ORA and GSEA/MSEA for ST000291 (enrichmet)

For ST000291 we can also build an `example_data` table required by the GSEA
part of `enrichmet()`:

```{r enrichmet-st291-example-data}
example_data_st291 <- st291_tt %>%
  filter(!is.na(HMDB)) %>%
  transmute(
    met_id = HMDB,
    pval   = !!rlang::sym("pvalue"),
    log2fc = !!rlang::sym("log2FC")
  )

cat("example_data_st291:\n")
print_tbl(head(example_data_st291))
```

Now we run `enrichmet()` requesting both ORA and GSEA:

```{r enrichmet-st291-run, eval=FALSE}
results_st291_enrichmet <- enrichmet(
  inputMetabolites     = st291_sig_HMDB,
  PathwayVsMetabolites = PathwayVsMetabolites_st291,
  example_data         = example_data_st291,
  top_n                = 100,
  p_value_cutoff       = 0.05,
  analysis_type        = c("enrichment", "gsea"),
  run_plots            = TRUE
)

st291_ora_enrichmet  <- results_st291_enrichmet$pathway_enrichment_results
st291_msea_enrichmet <- results_st291_enrichmet$gsea_results

cat("ORA (ST000291, enrichmet):\n")
print_tbl(head(st291_ora_enrichmet))

cat("\nMSEA/GSEA (ST000291, enrichmet):\n")
print_tbl(head(st291_msea_enrichmet))

# Optional plots:
# results_st291_enrichmet$pathway_plot
# results_st291_enrichmet$impact_plot
# results_st291_enrichmet$gsea_plot
```

# Enrichment with localEnrichment (skeleton)

This section assumes that `localEnrichment` provides helper functions such as
`filter_ES_by_background()`, `run_ORA()` and `run_MSEA()`, operating on
`EnrichmentSet` objects. The idea is to use the *same* HMDB signatures and
backgrounds as above.

```{r localEnrichment-skeleton, eval=FALSE}
library(localEnrichment)

# Construct EnrichmentSets restricted to the functional background
KEGG_cachexia_ES <- filterEnrichmentSet(KEGGset_global, ids = background_cachexia_HMDB)
KEGG_st291_ES    <- filterEnrichmentSet(KEGGset_global, ids = background_st291_HMDB)

# Example ORA for Cachexia
ora_cachexia_local <- run_ORA(
  sig = cachexia_sig_HMDB,
  bg  = background_cachexia_HMDB,
  ES  = KEGG_cachexia_ES
)

# Example ORA + MSEA for ST000291
ora_st291_local <- run_ORA(
  sig = st291_sig_HMDB,
  bg  = background_st291_HMDB,
  ES  = KEGG_st291_ES
)

msea_st291_local <- run_MSEA(
  ranked_stats = example_data_st291$log2fc,
  ids          = example_data_st291$met_id,
  ES           = KEGG_st291_ES
)

print_tbl(head(ora_cachexia_local))
print_tbl(head(ora_st291_local))
print_tbl(head(msea_st291_local))
```

> Adapt the calls above to match the exact function signatures of
> your `localEnrichment` implementation. The key idea is to reuse:
>
> - `cachexia_sig_HMDB`, `background_cachexia_HMDB`, `KEGG_cachexia_ES`
> - `st291_sig_HMDB`, `background_st291_HMDB`, `KEGG_st291_ES`

# ChemRich (outline)

ChemRich performs enrichment based on **chemical similarity clusters**, not
predefined pathways. It typically requires:

- effect size (e.g. log2FC),
- p-values,
- a structural identifier (InChIKey, SMILES, etc.).

Since ChemRich is provided as a web tool, the usual workflow is:

1. Prepare a CSV with at least ID, log2FC and p-value.
2. Upload to ChemRich web.
3. Export the results table.
4. Re-import into R for comparison.

For ST000291:

```{r chemrich-st291-export, eval=FALSE}
chemrich_input_st291 <- st291_tt %>%
  filter(!is.na(PubChemCID)) %>%
  transmute(
    PubChemCID = PubChemCID,
    log2FC     = log2FC,
    pvalue     = pvalue
  )

readr::write_csv(chemrich_input_st291,
                 "results/chemrich_ST000291_input.csv")
```

After running ChemRich on the web and exporting results:

```{r chemrich-st291-import, eval=FALSE}
chemrich_st291_results <- readr::read_csv("results/chemrich_ST000291_results.csv")
print_tbl(head(chemrich_st291_results))
```

# MetaboAnalyst Web (outline)

MetaboAnalyst Web is used here as a practical reference point (“gold
standard”). The workflow is:

1. Prepare an input table in the expected format (ID + intensities or statistics).
2. Upload and run:
   - ORA (e.g. KEGG pathways),
   - MSEA (for full tables such as ST000291).
3. Export results as CSV and save under `results/MetaboAnalyst/`.

```{r metaboanalyst-import, eval=FALSE}
ma_cachexia_ora <- readr::read_csv(
  "results/MetaboAnalyst/cachexia_kegg_ORA.csv"
)

ma_st291_ora <- readr::read_csv(
  "results/MetaboAnalyst/ST000291_kegg_ORA.csv"
)

ma_st291_msea <- readr::read_csv(
  "results/MetaboAnalyst/ST000291_kegg_MSEA.csv"
)

print_tbl(head(ma_cachexia_ora))
print_tbl(head(ma_st291_ora))
print_tbl(head(ma_st291_msea))
```

# Comparison of results (skeleton)

The comparison strategy is:

1. Harmonize pathway identifiers across tools (e.g. KEGG pathway IDs).
2. Select top-N hits per method (by adjusted p-value).
3. Build tables of overlaps and, optionally, visual summaries.

Below is a simple skeleton for comparing ORA results for ST000291 using KEGG.

```{r compare-ora-st291, eval=FALSE}
# Assume we have:
# - st291_ora_enrichmet  (from enrichmet)
# - ora_st291_local      (from localEnrichment)
# - ma_st291_ora         (from MetaboAnalyst)

clean_kegg <- function(x) {
  x %>%
    tolower() %>%
    str_trim()
}

# Adapt column names to your actual results
ora_st291_enrich <- st291_ora_enrichmet %>%
  mutate(source = "enrichmet",
         pathway_id = clean_kegg(Pathway))

ora_st291_local2 <- ora_st291_local %>%
  mutate(source = "localEnrichment",
         pathway_id = clean_kegg(pathway))

ora_st291_ma2 <- ma_st291_ora %>%
  mutate(source = "MetaboAnalyst",
         pathway_id = clean_kegg(Pathway))

ora_all <- bind_rows(
  ora_st291_enrich,
  ora_st291_local2,
  ora_st291_ma2
)

# Top 10 per method
ora_top10 <- ora_all %>%
  group_by(source) %>%
  slice_min(order_by = adj_pvalue, n = 10) %>%
  ungroup()

print_tbl(ora_top10, n = 30)
```

A similar approach can be used for MSEA results, focusing on ST000291 and
the methods that provide ranked enrichment (enrichmet, localEnrichment,
MetaboAnalyst, and ChemRich clusters in a more qualitative way).

---

This R Markdown document is meant to be executed **chunk by chunk** so that
you can inspect objects, adjust paths and function calls (especially those
depending on your localEnrichment implementation) and gradually build a
robust, reproducible benchmarking workflow.
