---
title: "ST000291 – POMA pipeline with HMDB subset"
author: "Alex Sánchez"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(tidyverse)
library(readr)
library(janitor)
library(limma)
library(POMA)
library(SummarizedExperiment)
library(rvest)
library(xml2)
```

# Introduction

This document contains a **clean, POMA-based preprocessing and analysis pipeline** for the Metabolomics Workbench dataset **ST000291**.

The goals are:

1. To obtain a **normalized, imputed feature matrix** using POMA.
2. To perform a **limma-based differential analysis** (Baseline vs Cranberry).
3. To construct **enrichment analysis inputs** (top table, selected list, background).
4. To build a **subset restricted to metabolites with HMDB IDs**, using the mapping file  
   `data/MW_ST000291__map.csv`.
5. To compare the full POMA-based analysis with the HMDB-only analysis.

We deliberately **avoid manual MWTab parsing** and rely instead on the higher-level
SummarizedExperiment objects returned by POMA, which ensures consistency with
previous analyses and facilitates comparison with MetaboAnalyst.

> **IMPORTANT**
>
> The original study followed a 2×2 crossover design, but the public ST000291
> dataset does **not** include subject identifiers. Therefore, we must treat
> samples as **independent** (unpaired) in the statistical analysis.  
> This increases variance and is conservative: it inflates the risk of false
> negatives rather than false positives.

# Data acquisition via POMA / Metabolomics Workbench API

We first obtain **SummarizedExperiment** objects for the negative and positive
modes using the `do_query()` function (provided by the Metabolomics Workbench
API client used in the original POMA-based workflow).

```{r get-sumexp}
# Negative mode
data_negative_mode <- do_query(
  context     = "study",
  input_item  = "analysis_id",
  input_value = "AN000465",
  output_item = "SummarizedExperiment"
)

# Positive mode
data_positive_mode <- do_query(
  context     = "study",
  input_item  = "analysis_id",
  input_value = "AN000464",
  output_item = "SummarizedExperiment"
)
```

Each SummarizedExperiment contains:

- an **assay matrix** (features × samples),
- **rowData** (feature annotations),
- **colData** (sample metadata, including treatment).

# Metabolite names and PubChem IDs

The metabolite identifiers are retrieved from Metabolomics Workbench as an HTML
table and then converted into a clean data frame with **Metabolite Name**,
**PubChem** and **KEGG** IDs. PubChem IDs will be used as **feature IDs**.

```{r metabolite-names}
metaboliteNamesURL <- "https://www.metabolomicsworkbench.org/data/show_metabolites_by_study.php?STUDY_ID=ST000291&SEARCH_TYPE=KNOWN&STUDY_TYPE=MS&RESULT_TYPE=1"

metaboliteNames_html <- metaboliteNamesURL %>%
  read_html() %>%
  html_nodes(".datatable")

metaboliteNames_negative <- metaboliteNames_html %>%
  .[[1]] %>%
  html_table() %>%
  dplyr::select(`Metabolite Name`, PubChemCompound_ID, `Kegg Id`)

metaboliteNames_positive <- metaboliteNames_html %>%
  .[[2]] %>%
  html_table() %>%
  dplyr::select(`Metabolite Name`, PubChemCompound_ID, `Kegg Id`)

metaboliteNames <- bind_rows(metaboliteNames_negative, metaboliteNames_positive) %>%
  dplyr::rename(
    names  = `Metabolite Name`,
    PubChem = PubChemCompound_ID,
    KEGG    = `Kegg Id`
  ) %>%
  mutate(
    KEGG    = ifelse(KEGG == "-", "UNKNOWN", KEGG),
    PubChem = ifelse(PubChem == "-", "UNKNOWN", PubChem)
  ) %>%
  filter(!duplicated(PubChem))

head(metaboliteNames)
```

# Build feature matrix and metadata

We construct a single feature matrix combining negative and positive mode,
then assign **PubChem IDs** as rownames.

```{r build-features}
## Negative mode features 
features_negative <- assay(data_negative_mode) %>%
  dplyr::slice(-n())  # remove last row if it corresponds to a non-feature

rownames(features_negative) <- rowData(data_negative_mode)$metabolite[1:(length(rowData(data_negative_mode)$metabolite)-1)]

## Positive mode features
features_positive <- assay(data_positive_mode) %>%
  dplyr::slice(-n())

rownames(features_positive) <- rowData(data_positive_mode)$metabolite[1:(length(rowData(data_positive_mode)$metabolite)-1)]

## Combine positive and negative mode and set PubChem IDs as feature names
features <- bind_rows(features_negative, features_positive) %>%
  tibble::rownames_to_column("names") %>%
  right_join(metaboliteNames, by = "names") %>%
  select(-names, -KEGG) %>%
  tibble::column_to_rownames("PubChem")

dim(features)
head(features[, 1:5])
```

Now we build the **sample metadata** (`pdata`) and ensure that the treatment
factor has the desired levels.

```{r metadata}
pdata <- colData(data_negative_mode) %>%   # or data_positive_mode; metadata are equal
  as.data.frame() %>%
  tibble::rownames_to_column("ID") %>%
  mutate(
    Treatment = case_when(
      Treatment == "Baseline urine"                        ~ "Baseline",
      Treatment == "Urine after drinking apple juice"      ~ "Apple",
      Treatment == "Urine after drinking cranberry juice"  ~ "Cranberry",
      TRUE ~ as.character(Treatment)
    )
  )

pdata$Treatment <- factor(pdata$Treatment,
                          levels = c("Baseline", "Apple", "Cranberry"))

table(pdata$Treatment)
```

# Create POMA object and preprocess (imputation + normalization)

We now create a POMA SummarizedExperiment object, then apply:

1. `PomaImpute()` – to handle missing values.
2. `PomaNorm()` – to normalize the data.

```{r poma-object}
data_sumexp <- PomaCreateObject(
  metadata = pdata,
  features = t(features)   # POMA expects features in columns
)

data_preprocessed <- data_sumexp %>%
  PomaImpute() %>%
  PomaNorm()

data_preprocessed
```

The normalized, imputed feature matrix is now:

```{r extract-preprocessed}
expr_all <- assay(data_preprocessed)              # rows: features (PubChem), cols: samples
sample_metadata <- colData(data_preprocessed) %>%
  as.data.frame()

dim(expr_all)
head(expr_all[, 1:5])
```

# Differential analysis with limma (full POMA data)

We focus on **Baseline vs Cranberry**, ignoring Apple for the main comparison.

```{r limma-full-setup}
group <- factor(sample_metadata$Treatment)

sel <- group %in% c("Baseline", "Cranberry")
expr_bc  <- expr_all[, sel]
group_bc <- droplevels(group[sel])
table(group_bc)
```

We fit a standard limma model:

```{r limma-full-fit}
design_full <- model.matrix(~ group_bc)
colnames(design_full)

fit_full <- lmFit(expr_bc, design_full)
fit_full <- eBayes(fit_full)

# coef = 2 corresponds to the contrast Cranberry vs Baseline
limma_full <- topTable(fit_full, coef = 2, number = Inf) %>%
  tibble::rownames_to_column("Feature")  # Feature = PubChem ID
```

We now add human-readable metabolite names by joining with `metaboliteNames`,
and also annotate with HMDB IDs using the MW mapping file in a later step.

```{r limma-full-annotate-basic}
metabolite_anno <- metaboliteNames %>%
  select(PubChem, Metabolite_name = names) %>%
  distinct()

limma_full <- limma_full %>%
  left_join(metabolite_anno, by = c("Feature" = "PubChem")) %>%
  relocate(Metabolite_name, .before = Feature)

head(limma_full, 10)
```

## Enrichment lists – full POMA data

For the full dataset, we define:

- **background_full_pubchem**: all features (PubChem IDs) tested in limma.
- **selected_full_pubchem**: features with FDR < 0.05.

```{r enrichment-lists-full}
background_full_pubchem <- limma_full$Feature

selected_full_pubchem <- limma_full %>%
  filter(adj.P.Val < 0.05) %>%
  pull(Feature)

length(background_full_pubchem)
length(selected_full_pubchem)
```

These objects can be used directly with enrichment tools that accept **PubChem**
IDs, or can be converted to HMDB using the mapping file.

# HMDB mapping and HMDB-based subset

We now load the file `data/MW_ST000291__map.csv`, which contains the mapping
from PubChem (as `Query`) to HMDB.

```{r hmdb-map}
MW_map <- read_csv("data/MW_ST000291__map.csv", show_col_types = FALSE) %>%
  clean_names()   # columns: query, match, hmdb, pubchem, chebi, kegg, metlin, smiles, comment

head(MW_map)
```

We build a simple table linking **PubChem IDs (as character)** to **HMDB IDs**.

```{r hmdb-map-prep}
map_pubchem_hmdb <- MW_map %>%
  mutate(PubChem = as.character(query)) %>%
  select(PubChem, HMDB = hmdb) %>%
  filter(!is.na(HMDB) & HMDB != "")

head(map_pubchem_hmdb)
```

We now annotate the full limma results with HMDB IDs and define HMDB-based
background and selected lists.

```{r limma-full-hmdb-annot}
limma_full <- limma_full %>%
  left_join(map_pubchem_hmdb, by = c("Feature" = "PubChem")) %>%
  relocate(HMDB, .after = Feature)

head(limma_full, 10)

# HMDB-based background and selected sets (from full POMA data)
background_full_hmdb <- limma_full %>%
  filter(!is.na(HMDB) & HMDB != "") %>%
  pull(HMDB) %>%
  unique()

selected_full_hmdb <- limma_full %>%
  filter(adj.P.Val < 0.05, !is.na(HMDB), HMDB != "") %>%
  pull(HMDB) %>%
  unique()

length(background_full_hmdb)
length(selected_full_hmdb)
```

At this point we have:

- A **full limma analysis** on normalized POMA data.
- PubChem-based and HMDB-based **background** and **selected** lists.

However, to mimic the situation where we **restrict the analysis beforehand**
to HMDB-annotated metabolites (as MetaboAnalyst would do), we also fit a
**separate limma model on an HMDB-only subset of features**.

# Limma analysis on HMDB-only subset

We first restrict the expression matrix to features with a valid HMDB ID
according to the mapping file.

```{r expr-hmdb-subset}
hmdb_pubchem_ids <- map_pubchem_hmdb$PubChem

expr_bc_hmdb <- expr_bc[rownames(expr_bc) %in% hmdb_pubchem_ids, , drop = FALSE]

dim(expr_bc_hmdb)
```

We then fit limma again on this restricted feature set.

```{r limma-hmdb-fit}
design_hmdb <- model.matrix(~ group_bc)

fit_hmdb <- lmFit(expr_bc_hmdb, design_hmdb)
fit_hmdb <- eBayes(fit_hmdb)

limma_hmdb <- topTable(fit_hmdb, coef = 2, number = Inf) %>%
  tibble::rownames_to_column("Feature")  # Feature = PubChem ID (subset)
```

We annotate again with metabolite names and HMDB IDs.

```{r limma-hmdb-annot}
limma_hmdb <- limma_hmdb %>%
  left_join(metabolite_anno, by = c("Feature" = "PubChem")) %>%
  left_join(map_pubchem_hmdb, by = c("Feature" = "PubChem")) %>%
  relocate(Metabolite_name, .before = Feature) %>%
  relocate(HMDB, .after = Feature)

head(limma_hmdb, 10)
```

Now we define the **HMDB-only** enrichment lists from this restricted analysis:

```{r enrichment-lists-hmdb}
background_hmdb <- limma_hmdb %>%
  filter(!is.na(HMDB) & HMDB != "") %>%
  pull(HMDB) %>%
  unique()

selected_hmdb <- limma_hmdb %>%
  filter(adj.P.Val < 0.05, !is.na(HMDB), HMDB != "") %>%
  pull(HMDB) %>%
  unique()

length(background_hmdb)
length(selected_hmdb)
```

These HMDB-based lists are precisely what we would feed to tools that require
HMDB identifiers (e.g. MetaboAnalyst, certain disease or pathway mappings, etc.).

# Comparison: full HMDB vs HMDB-subset analysis

To quantify the impact of restricting the analysis to HMDB-annotated features,
we compare the HMDB-based results from:

- The **full POMA analysis** (HMDB extracted a posteriori).
- The **HMDB-only limma analysis** (features restricted a priori).

```{r comparison-hmdb}
background_full_hmdb_ids <- background_full_hmdb
background_hmdb_ids      <- background_hmdb

selected_full_hmdb_ids <- selected_full_hmdb
selected_hmdb_ids      <- selected_hmdb

n_bg_full  <- length(background_full_hmdb_ids)
n_bg_hmdb  <- length(background_hmdb_ids)

n_sel_full <- length(selected_full_hmdb_ids)
n_sel_hmdb <- length(selected_hmdb_ids)

overlap_sel   <- length(intersect(selected_full_hmdb_ids, selected_hmdb_ids))
jaccard_sel   <- overlap_sel / length(union(selected_full_hmdb_ids, selected_hmdb_ids))

tibble(
  Metric = c("Background size (HMDB)", "Significant metabolites (HMDB)",
             "Overlap (significant HMDB)", "Jaccard index (significant HMDB)"),
  Full_POMA  = c(n_bg_full, n_sel_full, overlap_sel, jaccard_sel),
  HMDB_only  = c(n_bg_hmdb, n_sel_hmdb, NA, NA)
)
```

# Summary

In this document we have:

1. Built a **POMA-based pipeline** for ST000291 that:
   - retrieves SummarizedExperiment objects from Metabolomics Workbench,
   - constructs a combined feature matrix using PubChem IDs,
   - applies imputation and normalization via `PomaImpute()` and `PomaNorm()`.

2. Performed a **limma-based differential analysis** comparing Baseline vs
   Cranberry in the normalized POMA data, obtaining:
   - a **full top table** (`limma_full`),
   - **background** and **selected** sets based on PubChem and HMDB IDs.

3. Built a separate **HMDB-only limma analysis** by restricting the feature
   matrix to those metabolites with valid HMDB IDs using
   `data/MW_ST000291__map.csv`, providing:
   - HMDB-based **background** (`background_hmdb`),
   - HMDB-based **selected** list (`selected_hmdb`).

4. Quantified the impact of restricting to HMDB-annotated metabolites by
   comparing HMDB-based results extracted **a posteriori** from the full POMA
   analysis with those obtained from the HMDB-only limma fit.

These objects (top tables, selected lists, and backgrounds) can now be used
for downstream **enrichment analyses** with `localEnrichment`, MetaboAnalyst,
or other tools that accept PubChem and/or HMDB identifiers.
